---
title: "Writing Efficient Functions"
description: "Creating Efficient Custom Functions in R"
author:
  - name: Lucas Kantorowski
    url: https://lkantoro.github.io/PersonalWebsite/
    orcid: 0000-0002-5300-3075
    affiliation: Master of Science Student in Statistics @ California Polytechnic State University
    affiliation-url: https://www.calpoly.edu/ 
date: 06-08-2025
categories: [Quarto, R] # self-defined categories
citation: 
  url: https://lkantoro.github.io/PersonalWebsite/portfolio/2025-06-08-fourth-post/ 
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
---

Here are some custom functions made in R. The first are pulled [*R for Data Science*](https://r4ds.hadley.nz/).

Before I share my functions, I load in some necessary libraries.

```{r}
#| label: load_libraries
library(lubridate)
library(dplyr)
library('e1071')
library(tidyverse)
library(rlang)
```


**1** The `rescale01()` function below performs a min-max scaling to 
standardize a numeric vector, but infinite values are left unchanged. 

I rewrote `rescale01()` so that `-Inf` is mapped to 0, and `Inf` is mapped to 1.
I utilized `case_when()` to accomplish this.


```{r}
#| label: question-1


rescale01 <- function(x) {
  if (class(x) != "numeric") {
    stop("The provided vector is not numeric")
  }
  
  if (length(x) < 2) {
    stop("The provided vector is less than length 2")
  }
  
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  
  #making negative infinite and infinite values the min and max, respectively
  x = case_when(x == -Inf ~ rng[1], x == Inf ~ rng[2],.default = x)
  
  (x - rng[1]) / (rng[2] - rng[1])
  
  
}
```

```{r}
#| eval: false

#testing with NA values
rescale01(c(1,2,3,4,5,6,NA,-7,8,Inf, NA))

#testing with negative infinity and negative inifinity values
rescale01(c(1,2,3,-Inf,5,6,NA,-7,8,Inf, NA))

#testing with character values
rescale01(c(1,2,3,-Inf,5,'6',NA,-7,8,Inf, NA))

#tseting with vector of length 1
rescale01(c(1))

```






