---
title: "Writing Efficient Functions"
description: "Creating Efficient Custom Functions in R"
author:
  - name: Lucas Kantorowski
    url: https://lkantoro.github.io/PersonalWebsite/
    orcid: 0000-0002-5300-3075
    affiliation: Master of Science Student in Statistics @ California Polytechnic State University
    affiliation-url: https://www.calpoly.edu/ 
date: 06-08-2025
categories: [Quarto, R] # self-defined categories
citation: 
  url: https://lkantoro.github.io/PersonalWebsite/portfolio/2025-06-08-fourth-post/ 
draft: false # setting this to `true` will prevent your post from appearing on your listing page until you're ready!
---

Here are twelve custom functions made in R. The first ten are pulled [*R for Data Science*](https://r4ds.hadley.nz/). Each function is created and some common inputs are tested below.

Before I share my functions, I load in some necessary libraries.

```{r}
#| label: load_libraries
library(lubridate)
library(dplyr)
library('e1071')
library(tidyverse)
library(rlang)
```

**1** The `rescale01()` function below performs a min-max scaling to standardize a numeric vector, but infinite values are left unchanged.

I rewrote `rescale01()` so that `-Inf` is mapped to 0, and `Inf` is mapped to 1. I utilized `case_when()` to accomplish this.

```{r}
#| label: question-1


rescale01 <- function(x) {
  if (class(x) != "numeric") {
    stop("The provided vector is not numeric")
  }
  
  if (length(x) < 2) {
    stop("The provided vector is less than length 2")
  }
  
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  
  #making negative infinite and infinite values the min and max, respectively
  x = case_when(x == -Inf ~ rng[1], x == Inf ~ rng[2],.default = x)
  
  (x - rng[1]) / (rng[2] - rng[1])
  
  
}
```

```{r}
#| eval: false

#testing with NA values
rescale01(c(1,2,3,4,5,6,NA,-7,8,Inf, NA))

#testing with negative infinity and negative infinity values
rescale01(c(1,2,3,-Inf,5,6,NA,-7,8,Inf, NA))

#testing with character values
rescale01(c(1,2,3,-Inf,5,'6',NA,-7,8,Inf, NA))

#tseting with vector of length 1
rescale01(c(1))

```

**2** Here is a function that accepts a vector of birthdates and computes the age of each person in years.

```{r}
#| label: question-2

#using time_length function from lubridate package
#https://stackoverflow.com/questions/15569333/get-date-difference-in-years-floating-point

age_in_years = function(date_vect) {
  #accounting fro if birthday vector is already provided in date format
  if (class(date_vect) == "Date") {
    return(time_length(difftime(Sys.Date(), date_vect), "years"))
  }
  
  #calculate year difference between today and vector value
  return(time_length(difftime(Sys.Date(), as.Date(date_vect)), "years"))

}
```

```{r}
#| eval: false

#testing with expected input
age_in_years(c("2001-01-01", "2021-01-30", "2013-08-20"))


#testing with alternate date format
age_in_years(c("01/01/2001", "2021/01/30"))


#testing with alternate, unclear date format
age_in_years(c("01/30/2001", "2021/01/30"))


#testing with vector already in date format
age_in_years(as.Date(c("01/01/2001", "2021/01/30"))
)
```

**3** Here is a function that computes the variance and skewness of a numeric vector.

```{r}
#| label: question-3

variance_and_skewness = function(num_vec){
  
  if (class(num_vec) != "numeric") {
    stop("The provided vector is not numeric")
  }
  
  
  v_s_df = data.frame(Variance = var(num_vec, na.rm = TRUE), 
                      Skewness = skewness(num_vec, na.rm = TRUE))
  
  return(v_s_df)
}
```

```{r}
#| eval: false

#testing with normal vector
variance_and_skewness(c(1,2,3,4,5,7,8,89,9,87,0))

#testing with a character vector
variance_and_skewness(c(1,2,3,4,5,7,8,89,9,87,'0'))

#testing with a vector of length 1
variance_and_skewness(c(1))

#testing with NA values
variance_and_skewness(c(1,2,3,4,5,NA,8,89,9,87))

```

**4** Here, `both_na()` takes two vectors of the same length and returns the number of positions that have an `NA` in both vectors.

```{r}
both_na = function(vec1, vec2){
  
  #checking if columns are same length
  if (length(vec1) != length(vec2)) {
    stop("The provided vectors are not the same length")
  }
  
  
  #check for if a vector is non-numeric
  if (class(vec1) != "numeric" | (class(vec2) != "numeric"))  {
    stop("At least 1 provided vector is not numeric")
  }
  
  
  #adding up the NA status of column 1 and column 2, then filtering by col3 = 2
  return(data.frame(vec1, vec1, 
             row_num_nas = (is.na(vec1) + is.na(vec2))) %>% 
    filter(row_num_nas == 2) %>% #  filtering by is.na(col1) + is.na(col2) = 2
    summarise(N = n()) %>% as.numeric()) # counting number of columns with 2 NAs
  
  
}
#extract single value from 1x1 data frame: https://stackoverflow.com/questions/60864283/extract-the-single-value-from-a-1-x-1-data-frame-produced-with-dplyr-as-a-vector


```

```{r}
#| eval: false

#testing with basic input
vec_a = c(NA,3,4,6, NA, NA, NA, 11, NA)
vec_b = c(NA, NA, NA,3,4, 5,NA,3,5)
both_na(vec_a, vec_b)


#testing with 0 values
vec_a = c(1,2,3,4,5,6)
vec_b = c(64, 78, 456456, 88, 44, 29)
both_na(vec_a, vec_b)


#testing with 1 vector having character values
vec_a = c(NA,NA,3,4,5,6)
vec_b = c(NA, NA, NA, '2',3,4)
both_na(vec_a, vec_b)


#testing with different length vectors
vec_a = c(NA,3,4,5,6)
vec_b = c(NA, NA, NA, '2',3,4)
both_na(vec_a, vec_b)

```

## Data Frame Functions

These are functions that take data frames as inputs.

**5** This function replaces the NA values of a specified column with a value based on some function of the column. Users must provide the data frame, a numeric column of the data frame, and a function that works on numeric vectors.

```{r}
#| label: function-from-lab-2

impute_missing = function(df, ..., impute_fun = mean) {
  # Check to verify function input is a function
  if (!is.function(impute_fun)) {
    stop("The input for `impute_fun` must be a function.")
  }
  
  # Check for non-character columns (from live code functions.qmd)
  character <- df %>%
    select(...) %>%
    map_lgl(~ (is.character(.x) | is.factor(.x))) %>%
    keep(identity) %>%
    names()
  
  if (length(character) > 0) {
    stop("The following columns are characters: ",
         str_c(character, collapse = ", "))
  }

  newdf = df %>% 
    mutate(across(c(...), ~ replace_na(data = .x, 
                                       replace = impute_fun(.x, na.rm = TRUE))))
  return(newdf)

}
```

```{r}
#| eval: false

#testing with 1 numeric column
mean(nycflights13::flights$arr_delay, na.rm = TRUE)

test1 = impute_missing(nycflights13::flights, 
               arr_delay)

test1 %>% select(arr_delay) %>% group_by(arr_delay) %>% 
  summarise(N = n()) %>% arrange(desc(N)) %>% slice_head(n = 10)


#testing with non-default function
median(nycflights13::flights$arr_delay, na.rm = TRUE)

test2 = impute_missing(nycflights13::flights, 
               arr_delay, impute_fun = median)

test2 %>% select(arr_delay) %>% group_by(arr_delay) %>% 
  summarise(N = n()) %>% arrange(desc(N)) %>% slice_head(n = 10)


#testing with 2 numeric columns
test3 = impute_missing(nycflights13::flights, 
               arr_delay, dep_delay)

test3 %>% select(arr_delay) %>% group_by(arr_delay) %>% 
  summarise(N = n()) %>% arrange(desc(N)) %>% slice_head(n = 10)

test3 %>% select(dep_delay) %>% group_by(dep_delay) %>% 
  summarise(N = n()) %>% arrange(desc(N)) %>% slice_head(n = 10)


#testing with character column input
impute_missing(nycflights13::flights,carrier)

impute_missing(nycflights13::flights, 
               arr_delay, dep_delay, carrier)


#testing with non-real impute function
impute_missing(nycflights13::flights, dep_delay, impute_fun = apple)

```

The following function work with data similar to the `nycflights13` data.

**6** `filter_severe()` finds all flights that were cancelled or delayed by more than an hour.

```{r}
#| label: question-6

filter_severe = function(df = nycflights13::flights){
  
  #making sure dataset is similar to the flights data in format
  if ('dep_delay' %in% colnames(df) == FALSE | 
      'arr_time' %in% colnames(df) == FALSE){
    stop("Dataset provided does not have desired column names")
  }
  
  return(df %>% filter(is.na(arr_time) == TRUE | dep_delay > 60))
}
```

```{r}
#| eval: false

#testing for normal inputs
filter_severe(nycflights13::flights)


#testing for unexpected dataframe
filter_severe(mpg)

```

**7** `summarize_severe()` counts the number of cancelled flights and the number of flights delayed by more than an hour.

```{r}
#| label: question-7

summarize_severe = function(df = nycflights13::flights){
  #using previous function
  #counting number of cancelled flights
  #counting number of delayed flights by more than one hour
  
  return(filter_severe(df) %>% 
           summarise(Cancelled = sum(is.na(arr_time)),
                                         Delayed = sum(dep_time > 60, 
                                                       na.rm = TRUE)))
}
```

```{r}
#| eval: false

#testing expected input
summarize_severe(nycflights13::flights)


#testing function without necessary columns
summarize_severe(mpg)

```

**8** Here, `filter_severe()` allows users to supply the number of hours that should be used to filter the flights that were cancelled or delayed.

```{r}
#| label: question-8

filter_severe = function(df = nycflights13::flights, hours = 1){
  #making sure dataset is similar in format to flights dataset
  if ('dep_delay' %in% colnames(df) == FALSE | 
      'arr_time' %in% colnames(df) == FALSE){
    stop("Dataset provided does not have desired column names")
  }
  
  #making sure a numeric argument is used for hours
  if (class(hours) != 'numeric'){
    stop("Please provide a numeric qunaitity for delay hours")
  }
  
  return(df %>% filter(is.na(arr_time) == TRUE | dep_delay > hours * 60))
}
```

```{r}
#| eval: false

#testing for normal inputs
filter_severe(nycflights13::flights, 1)
filter_severe(nycflights13::flights, 10)

#testing for non-numeric delay
filter_severe(nycflights13::flights, '1')


#testing for unexpected dataframe
filter_severe(mpg, 4)

```

**9** The `summarize_weather()` function computes the minimum, mean, and maximum, of a user supplied variable.

```{r}
#| label: question-9

summarize_weather = function(df = nycflights13::flights, num_var){
  #make sure variable is numeric
  if (length(df %>% select({{num_var}}) %>% 
             purrr::discard(~!is.numeric(.))) == 0){
    stop('Please provide a numeric variable')
  }
  
  return(df %>% select({{num_var}}) %>%
            summarise(Minimum = min({{num_var}}, na.rm = TRUE),
            Mean = mean({{num_var}}, na.rm = TRUE),
            Maximum = max({{num_var}}, na.rm = TRUE)))
}  
```

```{r}
#| eval: false

#testing with normal input
summarize_weather(nycflights13::flights, dep_time)


#testing with non-numeric column
summarize_weather(nycflights13::flights, carrier)
```

**10** The `standardize_time()` function below converts a user supplied variable that uses clock time (e.g., `dep_time`, `arr_time`, etc.) into a decimal time (i.e. hours + (minutes / 60)).

```{r}
#| label: question-10

standardize_time = function(df = nycflights13::flights, time_var){
  #user supplied variable that uses clock time into decimal time
  if (englue('{{time_var}}') %in% 
      c('dep_time', 'sched_dep_time', 'arr_time', 'sched_arr_time') == FALSE){
    stop("Please provide a clock time variable to summarize")
  }
  
  return(df %>% 
    mutate({{time_var}} := floor({{time_var}}/100) * 60 + {{time_var}} %% 100))
}
```

```{r}
#| eval: false

#testing using expected input
standardize_time(nycflights13::flights, arr_time)
standardize_time(nycflights13::flights, sched_dep_time)

#testing supplying a non-time column
standardize_time(nycflights13::flights, day)
```

# Plotting Functions

**11** The `sorted_bars()` function below:

-   takes a data frame and a variable as inputs and returns a **vertical** bar chart
-   sorts the bars in decreasing order (largest to smallest)
-   adds a title that includes the context of the variable being plotted

```{r}
#| label: question-11

sorted_bars <- function(df, var) {
  
  #checking to make sure input variable is categorical
  if (length(df %>% select({{var}}) %>% purrr::discard(~is.numeric(.))) == 0){
    stop('Please provide a categorical variable')
  }
  
  #creating name object for label purposes
  name = englue("{{ var }}")
  
  return(df %>% select({{var}}) %>% mutate({{var}} := fct_infreq({{var}})) %>% 
    ggplot(aes(x = {{var}})) + 
    ggtitle(paste("Distribution of", name, "(decreasing)")) + 
    geom_bar())
  
}  
```

```{r}
#| eval: false

#keeping only numeric variables
#https://stackoverflow.com/questions/5863097/selecting-only-numeric-columns-from-a-data-frame

#testing using a expected input
sorted_bars(nycflights13::flights, carrier)
sorted_bars(mpg, trans)


#testing using a numeric variable
sorted_bars(nycflights13::flights, dep_delay)

```

# Iteration

**12:** Make a sorted barplot for every character variable in the `mpg` dataset (built into `ggplot2`).

```{r}
#| label: question-12

#could not get sorting to work with map function
unsorted_bars <- function(df, var) {
  
  #checking to make sure input variable is categorical
  if (length(df %>% select({{var}}) %>% purrr::discard(~is.numeric(.))) == 0){
    stop('Please provide a categorical variable')
  }
  
  #creating name object for label purposes
  name = englue("{{ var }}")
  
  return(df %>% select({{var}}) %>% 
    ggplot(aes(x = {{var}})) + 
    ggtitle(paste("Distribution of", name, "(decreasing)")) + 
    geom_bar())
  
}  
```

```{r}
#| eval: false

#barplots on all categorical variables of mpg
mpg %>% purrr::discard(~is.numeric(.)) %>% 
  names() %>% map(~unsorted_bars(df = mpg, var = .data[[.x]]))

#use of .data[[.x]]
#https://stackoverflow.com/questions/60305159/tunnel-data-variables-through-env-variable-with-operator

```

# Contributing to the R for Data Science Community!

The functions you wrote for exercises 1-10 came from *R for Data Science*. You could consider making a pull request to the repository for the solutions!

<https://github.com/mine-cetinkaya-rundel/r4ds-solutions>

To learn more about how to make a pull request I would suggest this article: <https://usethis.r-lib.org/articles/pr-functions.html>
